Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-25 article.tex
adjust global regularization and first part of appendix

\begin{abstract} In this work, we propose a simple, but yet efficient,
methodologically motivated method for the problem of connectome inference in
calcium imaging data. The proposed algorithm is made of two steps. First,
processing the raw signals to detect neural peak activities. Second, inferring
the degree of association between neurons from partial correlation statistics.
This paper summarizes the methodology that led us to win the Connectomics
Challenge, proposes a simplified version of our method and finally discusses
our results with respect to other inference methods.\end{abstract}

@@

\textcolor{red}{connection}Neurons communicate through short spikes, characterized by a high
frequency, while low frequencies of the signal correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i \label{eq:high-pass-filter},
\end{align}
shown in the plot of Figure \ref{fig:hp_curve}:

@@

The last step of our filtering procedure has for objective to decrease the importance of spikes that occur when there is a high global activity in the network with respect to  spikes that occur during normal activity. Indeed, we have conjectured that when a large part of the network is firing, the rate at which observations are made is not high enough for being able to detect interactions and, that it would therefore be preferrable to lower their importance by changing appropriatly their magnitude. Additionnally, it is well-known that neurons may also spike because of a global high activity (see \cite{stetter2012model}). In such  context, detecting pairwise neuron interactions from the firing activity is meaningless. 
Accordingly, we have chosen to apply to the signal outputted by $h$ the following function  
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}.
\end{align}
that has for effect to magnify the importance of spikes that occur in case of low global activity.

@@

>>>>

\begin{abstract} In this work, we propose a simple, but yet efficient, method for the problem of connectome inference in
calcium imaging data. The proposed algorithm is made of two steps. First,
processing the raw signals to detect neural peak activities. Second, inferring
the degree of association between neurons from partial correlation statistics.
This paper summarizes the methodology that led us to win the Connectomics
Challenge, proposes a simplified version of our method and finally discusses
our results with respect to other inference methods.\end{abstract}

@@

Furthermore, it can be observed that neurons communicate through short spikes, characterized by a high
frequency, while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i \label{eq:high-pass-filter},
\end{align}
shown in the plot of Figure \ref{fig:hp_curve}:

@@

The last step of our filtering procedure has for objective to decrease the importance of spikes that occur when there is a high global activity in the network with respect to  spikes that occur during normal activity. Indeed, we have conjectured that when a large part of the network is firing, the rate at which observations are made is not high enough for being able to detect interactions and, that it would therefore be preferrable to lower their importance by changing appropriatly their magnitude. Additionnally, it is well-known that neurons may also spike because of a global high activity (see \cite{stetter2012model}). In such  context, detecting pairwise neuron interactions from the firing activity is meaningless. 
Accordingly, we have chosen to apply to the signal outputted by $h$ the following function  
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}.
\end{align}
that has for effect to magnify the importance of spikes that occur in case of low global activity. Note that the particular case where $\sum_{j} x^{t}_j$ equals $0$ is solved by setting $w(x^{t}_i) = 1$.

@@

Sections~\ref{sec:filter}~to~\ref{sec:results} describe a simplified version
of the actual winning method. However, one could be interesting in the full
description. This appendix will therefore follow the progress of the main
paper highlighting optimizations.

Let us remark that parameter value choices were made to maximize ROC AUC score
on \textit{normal-1} and, to a lesser extent, \textit{normal-4}.

\subsection*{Signal processing}

As we introduce a composed function in section~\ref{sec:filter}, let us describe a new composed function with optimized filters to improve the quality of the inference process.

Besides filters
$f_1$ and $f_2$ respectively defined by equations \ref{eq:symetric-median} and \ref{eq:weighted-asymetric-median}, we also consider two other low-pass
filters:

\begin{align}
f_3(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i + x^{t+2}_i \label{eq:asymetric-median-forward}, \\
f_4(x^t_i) &= x^{t+3}_i + x^{t+2}_i + x^{t+1}_i + x_i^t.
\label{eq:asymetric-median}
\end{align}

Theses filters, even if individually they lead to less good performances of the inference algorithm than filters $f_1$ and
$f_2$, the averaging approach over these four low-pass
filters gets better.

As a computation trick, it appears that using a parametrized hard-threshold filter slighlty improves performances. Introducing the parameter $c$ in function $h$ yields to an optimized filter denoted as $h^*$
\begin{align}
h^*(x^t_i) = (x_i^t)^c \text{ } \mathbb{1}(x_i^t \ge \tau) \text{ with } \tau > 0,
\end{align}
where $c$ is set to $0.9$ after experiments.

The magnification can also be optimized by adjusting the effect differently with respect to the global activity. In that sense, we introduce the following function 
\begin{align}
 w^*(x^{t}_i) &= {(x^{t}_i + 1 )^{\left (1 + \frac{1}{\sum_{j} x^{t}_j}\right )}}^{m}
\end{align}
where $m$ depends on the low-pass filter and on the range of the global activity. We refer to our implemetation, available on ... , for the values of the parameters and especially the parameter $m$.

\subsection*{Averaging the connectome inference from partial correlation statistics}

\subsection*{Connectome inference from pairwise interaction statistics}

% NEXT APPENDIX SECTION 
% SUPPLEMENTARY RESULTS
% EXPLAIN EACH TAB AND VERIFY THE EXACTITUDE - RUN OPTIMIZED VERSION ON low-noise, ... ?
%

% EXPLAIN PCA - FIG

\vspace{5cm}
